---
title: "Part 2 Basic Inferential Data Analysis"
author: "Javier Chang"
date: "9/8/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2); library(gridExtra)
```

## Synopsis

In this part of the Course Project we will analyze the ToothGrowth data in the R datasets package according to the following instructions.

1. Load the ToothGrowth data and perform some basic exploratory data analyses
2. Provide a basic summary of the data.
3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose.
4. State conclusions and the assumptions needed for the conclusions.

## Data Source

We will load the ToothGrowth dataset from the built-in R data sets. Accordingly to R documentation ["ToothGrowth - The Effect Of Vitamin C On Tooth Growth In Guinea Pigs"](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/ToothGrowth), the response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice or ascorbic acid (a form of vitamin C and coded as VC).

```{r load_data}
library(datasets); data(ToothGrowth)
```

## Basic summary of the data

```{r summary}
summary(ToothGrowth)
```

## Tooth growth comparison by supp and dose

### Tooth growth comparison by supp for each dose

An unpaired two sample T-test is conducted to compare the two delivery methods (supp="OJ" or supp="VC) for each dose. The null hyphotesis is that there is not difference in tooth growth between the two delivery methods (supp). The T-test is run with confidence level of 95% and the assumption that the variances are unequal for each dose.

```{r comparison by supp}
results <- data.frame(matrix(ncol = 4, nrow = 0))
for (idose in c(0.5, 1.0, 2.0)){
      ttest<- t.test(ToothGrowth$len[ToothGrowth$dose==idose & ToothGrowth$supp=="OJ"],
                  ToothGrowth$len[ToothGrowth$dose==idose & ToothGrowth$supp=="VC"], 
                  paired=FALSE, alternative = "two.sided", var.equal=FALSE, conf.level=0.95)
      results <- rbind(results, c(idose, ttest$p.value, ttest$conf.int))}
colnames(results) <- c("dose", "p.value", "conf.int.min", "conf.int.max")
knitr::kable(results, caption="Unpaired two sample T-test (supp OJ vs VC)",
             format.args = list(nsmall=4, digits=2))
```

As we can see from the previous table, the p-values are very low (<0.007) for the doses of 0.5 and 1 mg/d supporting the alternative hypothesis that there is a statistical difference of tooth length between supp OV and VC for those doses. However, the p-value is very high (0.96) for the 2.0 mg/d dose, supporting the null hypothesis that there is no statistical difference between the supp OV and VC in that dose.

### Tooth growth comparison by dose for each supp

Now we will compare tooth growth by each combination of doses for each supp. The null hypothesis is that there is no difference between them on each combination of groups. We will use an unpaired two sample T-test with confidence level of 95% and the assumption that the variances are unequal. The results of each combination of groups are shown in the following table.

```{r comparison by dose}
results <- matrix(ncol = 5, nrow = 0)
combinations <- matrix(c(2, 1, 2, 1, 0.5, 0.5), ncol=2, nrow=3);
colnames(combinations) =c("dose1", "dose2")
for (isupp in levels(ToothGrowth$supp)){
      for (i in 1:3){
      ttest<- t.test(ToothGrowth$len[ToothGrowth$dose==combinations[i, "dose1"] &
                                           ToothGrowth$supp==isupp],
                  ToothGrowth$len[ToothGrowth$dose==combinations[i, "dose2"] &
                                        ToothGrowth$supp==isupp], 
                  paired=FALSE, alternative = "two.sided", var.equal=FALSE, conf.level=0.95)
      results <- rbind(results, c(as.character(isupp), paste(combinations[i,"dose1"], "vs", 
                  combinations[i,"dose2"], sep=" "), round(ttest$p.value, 6),
                  round(ttest$conf.int, 4)))
      }}
colnames(results) <- c("supp", "dose", "p.value", "conf.int.min", "conf.int.max")
knitr::kable(data.frame(results), caption="Unpaired two sample T-test by dose",
             format.args = list(nsmall=4, digits=2))
```

As we can see in the previous table, the p-values are all less than 0.04, so there is a statistically significant difference between each group. Therefore, it can be stated that the higher the dose used, greater growth is observed in the teeth according to each supp

## Conclusions

Based on the analyzed data we can conclude the following

* The higher the dose, the greater the tooth growth regardless of the supplement method
* With orange juise (OJ) supplement method, greater tooth growth is obtained than with vitamin C (VC) supplement method for doses 0.5 mg/d and 1.0 mg/d. However at 2.0 mg/d dose no significant difference between OJ and VC methods.

Assumptions

* The groups of measurements are not paired and their variances are not equal
* The measurements are independent and they were selected at random

## Appendix

### Tooth growth by supp and dose

```{r plot tooth growth, fig.height=2, fig.width=4}
ggplot(ToothGrowth, aes(x=dose, y=len) )+
        ggtitle("ToothGrowth by dose and supp")+ theme(plot.title = element_text(hjust = 0.5))+
        geom_point(size=0.75, aes(color=supp)) + facet_wrap(~supp, ncol=2)   
```

### Code for this report

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```